Liveness-Based Garbage Collection for Lazy Languages

### 2015-11-27 paper/paper.tex

\begin{abstract} 
We consider  the problem of reducing  the memory required to  run lazy
first-order functional programs. We  address the problem by analyzing
programs  for liveness  of  heap-allocated data.   The  result of  the
analysis is  used to preserve  only live data---a subset  of reachable
data---during garbage  collection.  The result  is an increase  in the
garbage reclaimed and  a  reduction in  the peak memory requirement 
of programs.   While this  technique has already  been shown  to yield
benefits for  eager first-order  languages, the  lack of  a statically
determinable execution  order and  the presence  of closures  pose new
challenges  for lazy  languages.  These  requires changes  both in  the
liveness analysis itself and in the design of the garbage collector.

%% We use the  result of the analysis to annotate  each potential garbage
%% collection  point  in   the  program  with  a   set  of  deterministic
%% finite-state automata (DFA) describing the liveness at that point.
We
implement a copying collector that uses the results of the liveness
analysis to preserve live
objects, both evaluated (i.e.  in WHNF) and closures.  Our experiments
confirm  that  for  programs  running with  a  liveness-based  garbage
collector,   there  is   a   significant  decrease   in  peak   memory
requirements.   In  addition,   a  sizable  reduction  in   the  number  of
collections  ensures that  in spite  of using  a more  complex garbage
collector,  the execution  time  of  a program  remains comparable  with
the same program running with a reachabilty-based collector.
\warning{On our benchmark programs, we obtain a decrease of up to 75\%
  in the number of collections, and up to 50\% decrease in peak memory
  requirements.}
\end{abstract}

@@

In this paper,  we propose the use of liveness  analysis of heap cells
for garbage collection in a lazy first-order functional language.  The
central notion in our analysis  is a generalization of liveness called
{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
expression.   The  analysis has  two  parts.   First, we  calculate  a
context-sensitive  summary   of  each   function  as  a   {\em  demand
  transformer} that transforms a {\em  symbolic} demand on its body to
demands on its  arguments.  This summary is used  to step through
function  calls during  analysis.  Second,  the concrete  demand on  a
function   body  is   obtained  through   a  0-CFA-like   conservative
approximation  that combines  the  demands  on all  the  calls to  the
function.   The result  of the  analysis is  an annotation  of certain
program points  with deterministic finite-state  automata (DFA)  capturing the  liveness of
variables  at  these points.  Depending  on  the program  point  where
garbage  collection has  been  triggered,  the garbage collector consults  a set  of
automata to restrict reachability during  marking.  This results in an
increase in  the garbage reclaimed  and consequently in  fewer garbage
collections.

While this idea has been shown to be effective for a
first-order {\em  eager} language~\cite{asati14lgc}, a straightforward  extension of the
technique   is  not   possible  for  lazy   languages,  where
heap-allocated objects  may include closures  (runtime representations
of unevaluated expressions). Since data  is made live by evaluation of
closures, and  in lazy languages the  place in the program  where this
evaluation  takes  place  cannot be  statically  determined,  laziness
complicates  liveness  analysis  itself.    In  addition,  apart  from
evaluated   data,  we  need to  extend   liveness-based  collection   to
closures i.e. we reclaim the space for closures  which have no future  use.
Since a closure can  escape the scope in which it  was created, it has
to carry the liveness information  of its free variables. Moreover, as
execution  progresses and  possible  future uses  are eliminated,  the
liveness information is updated with a more precise version.  All this
requires significant  modification to  the earlier  garbage collection
scheme for eager languages.

@@

\newcommand{\emm}[2]{\ensuremath{\mathcal{#1}_{#2}}}
\subsection{Generating Equations for \Lfonly$_\mathit{f}$}
Given a  function $\mathit{f}$, we now  describe how to
generate  equations   for  the   demand  transformation
\Lfonly$_\mathit{f}$.        The       program       in
Figure~\ref{fig:mot-example2}   serves  as   a  running
example.  Starting with a  symbolic demand $\sigma$, we
determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   In
particular, we  consider $\Lanv{1}{x_i}$,  the liveness
of the $i^{\mbox{\footnotesize th}}$ parameter $x_i$ at
the program point at  the beginning of $e_{\mathit{f}}$
(assumed   to   be   $\pi_1$).   By   the   rule   {\sc
  live-define},   this   should    be   the   same   as
$\Lf{f}{i}{\sigma}$. Applying this to \length, we have:

\begin{eqnarray*}
&& \Lanv{1}{\xl} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
  \cup \clazy\Lf{\length}{1}{\clazy\sigma}
\end{eqnarray*}
%% and  the only equation defining \Lfone{\length} is:
%%   \begin{eqnarray*}
%%    && \Lf{\length}{1}{\sigma}
%%     =  \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma} \cup
%% \clazy\Lf{\length}{1}{\clazy\sigma}
%% \end{eqnarray*}
and  the equation defining \Lfone{\fun} is:
\begin{eqnarray*}
  && \Lf{\fun}{1}{\sigma}
  =  \clazy\sigma \cup \Lf{\length}{1}{\bcar\sigma_{\mathit{\length}}} \cup  \Lf{\length}{1}{\clazy\sigma} 
  \cup  \acdr\Lf{\fun}{1}{\bcdr\sigma_{\mathit{\fun}}}
\end{eqnarray*}

In general, the equations for \Lfonly\ are recursive as
in the  case of  \Lfonly$_\mathit{f}$. As  mentioned in
\cite{asati14lgc},   a   closed   form   solution   for
$\Lfone{\mathit{f}}$ can  be derived by  observing that
each of  the liveness rules  modifies a demand  only by
prefixing  it with  symbols  in  the alphabet  $\lbrace
\acar, \acdr,\bcar,  \bcdr, \clazy  \rbrace$. Therefore
we can  assume that $\Lf{f}{i}{\sigma}$ has  the closed
form:
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} = \Df{f}{i}\sigma
\end{eqnarray}
where \Df{f}{i}  are sets of strings  over the alphabet
mentioned above.  Substituting the  guessed form in the
equation   describing    \Lfonly$_{\mathit   f}$,   and
factoring  out   $\sigma$,  we  get  an   equation  for
\Df{f}{i}  that   is  independent  of   $\sigma$.   Any
solution   for   \Df{f}{i}   yields  a   solution   for
\Lfonly$_{\mathit  f}$.   Applied to  \Lfonly$_{\mathit
  \length}$, we get:
  \begin{eqnarray*}
&&  \Lf{\fun }{1}{\sigma} = \Df{\fun}{1}\sigma,~\mbox{and}\\
&&   \Df{\fun}{1} =  \clazy \cup \Df{\length}{1}\bcar \cup  \Df{\length}{1}\clazy 
       \cup  \acdr\Df{\fun}{1}\bcdr
  \end{eqnarray*}

@@

%% For  the  running  example,   $\length$  has  calls  from  $\main$  at
%% $\pi_{12}$ and a recursive call at $\pi_6$.
%% So $\sigma_{\length} =
%%      \deltacall{\length}{\pi_{12}}{\main}  \cup
%% \deltacall{\length}{\pi_6}{\length}$.
Filling in  the values gives:
\begin{eqnarray*}
\sigma_{\fun}    &=&
 \{\sigma_{\mathit{all}}\}  ~\cup~{\bcdr\sigma_{\mathit{\fun}}}
\end{eqnarray*}
As examples, the liveness of \Lanv{2}{\pl} and \Lanv{1}{\py} in terms
of  $\sigma_{\fun}$ are:
\begin{align*}
\Lanv{2}{\pl} &= \clazy \cup \acar\Lf{\fun}{2}{\bcdr\sigma_{\mathit{\fun}}}
 \\
\Lanv{1}{\py} &= \clazy \cup \bcar\sigma_{\mathit{\fun}}
\end{align*}

>>>>

\begin{abstract} 
We consider  the problem of reducing  the memory required to  run lazy
first-order functional programs. Our approach is to analyze
programs  for liveness  of  heap-allocated data.   The  result of  the
analysis is  used to preserve  only live data---a subset  of reachable
data---during garbage  collection.  The result  is an increase  in the
garbage reclaimed and  a  reduction in  the peak memory requirement 
of programs.   While this  technique has already  been shown  to yield
benefits for  eager first-order  languages, the  lack of  a statically
determinable execution  order and  the presence  of closures  pose new
challenges  for lazy  languages.  These  requires changes  both in  the
liveness analysis itself and in the design of the garbage collector.

%% We use the  result of the analysis to annotate  each potential garbage
%% collection  point  in   the  program  with  a   set  of  deterministic
%% finite-state automata (DFA) describing the liveness at that point.
To show the effectiveness of our method, we
implement a copying collector that uses the results of the liveness
analysis to preserve live 
objects, both evaluated (i.e.  in WHNF) and closures.  Our experiments
confirm  that  for  programs  running with  a  liveness-based  garbage
collector,   there  is   a   significant  decrease   in  peak   memory
requirements.   In  addition,   a  sizable  reduction  in   the  number  of
collections  ensures that  in spite  of using  a more  complex garbage
collector,  the execution  time  of  a program  remains comparable  with
the same program running with a reachabilty-based collector.
\warning{On our benchmark programs, we obtain a decrease of up to 75\%
  in the number of collections, and up to 50\% decrease in peak memory
  requirements.}
\end{abstract}

@@

In this paper,  we propose the use of liveness  analysis of heap cells
for garbage collection in a lazy first-order functional language.  The
central notion in our analysis  is a generalization of liveness called
{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
expression.   The  analysis has  two  parts.   First, we  calculate  a
context-sensitive  summary   of  each   function  as  a   {\em  demand
  transformer} that transforms a {\em  symbolic} demand on its body to
demands  on its  arguments.   This  summary is  used  to step  through
function  calls during  analysis.  Second,  the concrete  demand on  a
function  body is  obtained through  a conservative approximation
similar to 0-CFA~\cite{Shivers:1988} that combines the  demands on all the calls
to  the function.   The result  of the  analysis is  an annotation  of
certain program points with  deterministic finite-state automata (DFA)
capturing the liveness of variables at these points.  Depending on the
program point where garbage collection has been triggered, the garbage
collector consults a  set of automata to  restrict reachability during
marking.  This  results in  an increase in  the garbage  reclaimed and
consequently in fewer garbage collections.

While this idea has been shown to be effective for a
first-order {\em  eager} language~\cite{asati14lgc}, a straightforward  extension of the
technique   is  not   possible  for  lazy   languages,  where
heap-allocated objects  may include closures  (runtime representations
of unevaluated expressions). Since data  is made live by evaluation of
closures, and  in lazy languages the  place in the program  where this
evaluation  takes  place  cannot be  statically  determined,  laziness
complicates  liveness  analysis  itself.    In  addition, laziness
also complicates garbage collection, since, 
 apart  from
evaluated   data,  we  need to extend   liveness-based  collection   to
closures.
Since a closure can  escape the scope in which it  was created, it has
to carry the liveness information  of its free variables. Moreover, as
execution  progresses and  possible  future uses  are eliminated,  the
liveness information is updated with more precise versions.  All this
requires significant  modification to  the earlier  garbage collection
scheme for eager languages.

@@

\newcommand{\emm}[2]{\ensuremath{\mathcal{#1}_{#2}}}

\subsection{Generating Equations for \Lfonly$_\mathit{f}$}
Given a  function $\mathit{f}$, we now  describe how to
generate  equations   for  the   demand  transformation
\Lfonly$_\mathit{f}$.        The       program       in
Figure~\ref{fig:mot-example2}   serves  as   a  running
example.  Starting with a  symbolic demand $\sigma$, we
determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   In
particular, we  consider $\Lanv{1}{x_i}$,  the liveness
of the $i^{\mbox{\footnotesize th}}$ parameter $x_i$ at
the program point at  the beginning of $e_{\mathit{f}}$
(assumed   to   be   $\pi_1$).   By   the   rule   {\sc
  live-define},   this   should    be   the   same   as
$\Lf{f}{i}{\sigma}$. Applying this to \length, we have:

\begin{eqnarray*}
&& \Lanv{1}{\xl} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
  \cup \clazy\Lf{\length}{1}{\clazy\sigma}
\end{eqnarray*}

%% and  the only equation defining \Lfone{\length} is:
%%   \begin{eqnarray*}
%%    && \Lf{\length}{1}{\sigma}
%%     =  \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma} \cup
%% \clazy\Lf{\length}{1}{\clazy\sigma}
%% \end{eqnarray*}
and  the equation defining \Lfone{\fun} is:
\begin{eqnarray*}
  && \Lf{\fun}{1}{\sigma}
  =  \clazy\sigma \cup \Lf{\length}{1}{\bcar\sigma_{\mathit{\length}}} \cup  \Lf{\length}{1}{\clazy\sigma} 
  \cup  \acdr\Lf{\fun}{1}{\bcdr\sigma_{\mathit{\fun}}}
\end{eqnarray*}

In general, the equations for \Lfonly\ are recursive as
in the  case of  \Lfonly$_\mathit{f}$. As  mentioned in
\cite{asati14lgc},   a   closed   form   solution   for
$\Lfone{\mathit{f}}$ can  be derived by  observing that
each of  the liveness rules  modifies a demand  only by
prefixing  it with  symbols  in  the alphabet  $\lbrace
\acar, \acdr,\bcar,  \bcdr, \clazy  \rbrace$. Therefore
we can  assume that $\Lf{f}{i}{\sigma}$ has  the closed
form:
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} = \Df{f}{i}\sigma
\end{eqnarray}

where \Df{f}{i}  are sets of strings  over the alphabet
mentioned above.  Substituting the  guessed form in the
equation   describing    \Lfonly$_{\mathit   f}$,   and
factoring  out   $\sigma$,  we  get  an   equation  for
\Df{f}{i}  that   is  independent  of   $\sigma$.   Any
solution   for   \Df{f}{i}   yields  a   solution   for
\Lfonly$_{\mathit  f}$.   Applied to  \Lfonly$_{\mathit
  \length}$, we get:
  \begin{eqnarray*}
&&  \Lf{\fun }{1}{\sigma} = \Df{\fun}{1}\sigma,~\mbox{and}\\
&&   \Df{\fun}{1} =  \clazy \cup \Df{\length}{1}\bcar \cup  \Df{\length}{1}\clazy 
       \cup  \acdr\Df{\fun}{1}\bcdr
  \end{eqnarray*}

@@

%% For  the  running  example,   $\length$  has  calls  from  $\main$  at
%% $\pi_{12}$ and a recursive call at $\pi_6$.
%% So $\sigma_{\length} =
%%      \deltacall{\length}{\pi_{12}}{\main}  \cup
%% \deltacall{\length}{\pi_6}{\length}$.

Filling in  the values gives:
\begin{eqnarray*}
\sigma_{\fun}    &=&
 \{\sigma_{\mathit{all}}\}  ~\cup~{\bcdr\sigma_{\mathit{\fun}}}
\end{eqnarray*}
As examples, the liveness of \Lanv{2}{\pl} and \Lanv{1}{\py} in terms
of  $\sigma_{\fun}$ are:
\begin{align*}
\Lanv{2}{\pl} &= \clazy \cup \acar\Lf{\fun}{2}{\bcdr\sigma_{\mathit{\fun}}}
 \\
\Lanv{1}{\py} &= \clazy \cup \bcar\sigma_{\mathit{\fun}}
\end{align*}
